\MakeUppercase{Linear versus centred chromatic numbers}\thanks{This research was partly funded by NSERC.}

### 2024-04-04 referee-response.md

With the current title a reader will still learn, within the first two pages that the improvement comes only by studying the linear chromatic number pseudogrids and decide for themselves whether it's worthwhile reading the rest of the argument.  (These first two pages also have the benefit of catching-up a reader on the current state of the art and the limitations of Kun et al's argument.)

**Referee Comment 2:** There are a few places (e.g. in section 3.6 -- see PDF) where the authors cut a few corners in their proofs; this should be addressed, but I don't think it is anything serious.

We have gone through the attached PDF carefully and corrected the minor issues pointed out. Some larger changes in response to the referee's comments are discussed below.

**Referee Comment 3:** A lot of the arguments in the middle sections, such as the packing lemma, seem to use the underlying intuition that grids (and by extension pseudogrids) look like the plane with the taxicab metric and the Lebesgue measure. I imagine the authors have already attempted this, but is there really no way to make that intuition rigorous by defining a metric and/or a measure on appropriate sets, and express the lemmas in terms of those? Something like that would make arguments such as the one in Lemma 13 much cleaner. In this direction, have the authors considered the idea of using a grid with every edge subdivided once as the index set for the path partition of the pseudogrid (the subdivision vertices symbolising the edges), with the purpose of making definitions such as the r-boxes more uniform (and of having a symmetric relation, whereby if $\mu$ is in the r-box of $\nu$, then $\nu$ should be in the r-box of $\mu$)?

### 2024-04-04 referee-response.md

We thank the referee for carefully reading the manuscript and giving thoughtful and helpful feedback.  The referee had 4 broad comments:

@@

**Referee Comment 1:** The paper's title is a bit ambitious since, as the authors point out, the machinery for relating the linear chromatic number to the centred one was already introduced previously (and this manuscript does not develop on it at all). A more honest title would be something along the lines of "The linear chromatic number of pseudogrids".

**Response:** We have chosen to leave the title as is.  In our opinion, the current title doesn't promise much.  The proposed title, on the other hand, suggests that the results apply only to pseudogrids.  Consider, for example, the similar-sounding title "The linear chromatic number of cactus graphs."

With the current title a reader will still learn, within the first two pages, that the improvement comes only by studying the linear chromatic number pseudogrids and decide for themselves whether it's worthwhile reading the rest of the argument.  (These first two pages also have the benefit of catching-up a reader on the current state of the art and the limitations of Kun et al's argument.)

@@

**Referee Comment 3:** A lot of the arguments in the middle sections, such as the packing lemma, seem to use the underlying intuition that grids (and by extension pseudogrids) look like the plane with the taxicab metric and the Lebesgue measure. I imagine the authors have already attempted this, but is there really no way to make that intuition rigorous by defining a metric and/or a measure on appropriate sets, and express the lemmas in terms of those? Something like that would make arguments such as the one in Lemma 13 much cleaner. In this direction, have the authors considered the idea of using a grid with every edge subdivided once as the index set for the path partition of the pseudogrid (the subdivision vertices symbolising the edges), with the purpose of making definitions such as the r-boxes more uniform (and of having a symmetric relation, whereby if $\mu$ is in the r-box of $\nu$, then $\nu$ should be in the r-box of $\mu$)?

During the initial writing phase of this result, we tried several such strategies.  The main annoying issue is that mapping the vertices of a subdivided grid onto objects in the pseudogrid sometimes maps a vertex onto a single edge of the pseudogrid (but not the endpoints of that edge).  In terms of colouring, this means that the subdivision vertex has an empty colour set.  It is possible to define things carefully enough for this to work, but then the reader is constantly having to verify that simple-sounding statements are correct under these careful definitions.

**Referee Comment 4:** In the acknowledgements, the authors mention that it was pointed out to them the problem on a grid (as opposed to pseudogrid) has a simple solution. Assuming the person who pointed that out is happy with it, it might be interesting to include the solution.

**Response:** We have added the solution along with the acknowledgement.

@@

**Referee Comment:** Top of Page 3: This somewhat addresses the previous comment.

**Response:** We have added further explanation (at the location of the second comment) of the limits of this argument here. In particular, even with an Excluded Grid Theorem that matches the currently known lower bound, this would only establish that the centred chromatic number is at most cubic in the linear chromatic number.  Even for planar graphs, the argument can only show that the centred chromatic number is at most quadratic in the linear chromatic number.

**Referee Comment:** I do not understand how this works when removing edges in $P_{v_i}$ -- doesn't $P_{v_i}$ belong to both a row and a column (and thus deleting it leaves us with both fewer rows and columns?)

**Response:** The referee is correct.  The description of deletion has been corrected.  Now we only delete edges of $G$ that correspond to edges of the grid (and then repeatedly remove vertices of degree at most $1$).  This avoids disconnecting a column when deleting a row, and vice versa.

### 2024-04-04 lin-vs-cen.tex
Addressed everything the referee asked. Just need to reread.

\documentclass{patmorin}
\listfiles
\usepackage{pat}
\usepackage{paralist}
\usepackage{dsfont}  % for \mathds{A}
\usepackage[utf8x]{inputenc}
\usepackage{skull}
\usepackage{paralist}
\usepackage{graphicx}
\usepackage[noend]{algorithmic}
\usepackage{bbm}  % needed for \mathbbm{1}
\usepackage{listings}

@@

The main subject of this paper was posed as an open problem at the 9th Annual Bellairs Workshop on Geometry and Graphs (v2), which was held January 21--28, 2022 and again at the 16th Annual Workshop on Probability and Combinatorics, which was held March 25--April 1, 2022. We are grateful to the organizers and participants of both workshops for providing a stimulating research environment.
Pat~Morin especially enjoyed early-morning discussions with Stefan~Langerman on the Seabourne Terrace. We are also grateful to Omer Angel for pointing out that the problem on a $k\times k$ grid (as opposed to pseudogrid) has a simple solution. (By partitioning the rows into consecutive pairs and the columns  into consecutive pairs, and then repeatedly deleting a pair of rows and columns whose intersection contains a vertex whose colour is unique, one is left with a graph that has a Hamiltonian path in which each colour apppears on at least two vertices.)

>>>>

\documentclass{patmorin}
\listfiles
\usepackage{pat}
\usepackage{paralist}
\usepackage{dsfont}  % for \mathds{A}
\usepackage[OT1]{fontenc}
\usepackage[utf8x]{inputenc}
\usepackage{skull}
\usepackage{paralist}
\usepackage{graphicx}
\usepackage[noend]{algorithmic}
\usepackage{bbm}  % needed for \mathbbm{1}
\usepackage{listings}

@@

The main subject of this paper was posed as an open problem at the 9th Annual Bellairs Workshop on Geometry and Graphs (v2), which was held January 21--28, 2022 and again at the 16th Annual Workshop on Probability and Combinatorics, which was held March 25--April 1, 2022. We are grateful to the organizers and participants of both workshops for providing a stimulating research environment.
Pat~Morin especially enjoyed early-morning discussions with Stefan~Langerman on the Seabourne Terrace.

We are also grateful to Omer Angel for pointing out that the problem on a $k\times k$ grid (as opposed to pseudogrid) has a simple solution: Partition the grid into $2\times 2$ grids and repeatedly delete a subgrid if it contains a vertex whose colour is unique.  If this causes the grid to become disconnected, keep only the largest remaining component and continue.  This process deletes at most $4\epsilon k$ vertices.  For $\epsilon < 1/4$, this implies that the graph that remains at the end has at least $k^2/2$ vertices.  In particular, the graph that remains is non-empty and every colour that appears in this graph appears in at least two vertices.  Finally, this graph has a Hamiltonian path (which can be constructed by traversing the boundary of  a spanning-tree of the dual graph).
